Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-02-16 SIR_manuscript.tex
more David's comments were addresses, change of subscr to roman in \_

\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\mathcal{R}_0}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\comment}{\showcomment}
\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

@@

We developed a deterministic model, Eqs. \eqref{model}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of the basic reproduction number $\Rnum$ are presented in the appendixs.

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=0.3]{./pix/sir_comp.png}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. For the description of the parameters see \cref{tab:params}.
\label{fig:flowchart}}
\end{center} 
\end{figure}

Table~\ref{tab:params} defines the model parameters, which are generally straightforward
\percap flows between compartments, or modifiers to these flow rates.
\david{you need to define what you mean by ``testing intensity'' $\rho$}
The novel component of the model comes in through the compartment-specific relative testing weights $W_S$, $W_I$ and $W_R$; these give the relative rates at which people in the $S$, $I$, and $R$ compartments are tested, respectively. For example, $W_I/W_S=2$ means that infected individuals are tested at twice the \percap rate of susceptible individuals.

In order to link to more applied models\david{what does that mean?  what is an ``applied model''?}, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S_u + W_I I_u + W_R R_u$, and calculating a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, the \percap testing rate for compartment $X$ is $F_X=\sigma W_X$, where $X \in \{S,I,R\}$. 
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I_c$, $R_c$) compartments and are thus unavailable for further testing.  Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can (if large enough) exhaust the pool of people available for testing, leading to a singularity when no one is left untested.  Although this phenomenon does not affect our analysis of $\Rnum$, it can affect the temporal dynamics (we discuss an adjustment to the model that solves this problem in the appendix).

\david{A comment on notation:  You haven't actually defined $S_u$, $I_u$ etc.  In any case, I urge you to use macros and use Roman font for subscripts that are textual as opposed to variables.  So $S_{\rm u}$ and $I_{\rm p}$ etc.  Then it is clear the subscript isn't a variable with a value.}\ali{will do}

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population\david{What's the difference btw ``well-mixed'' and ``homoegeneous''?}\ali{ non structured eg age, ok?}, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection, $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I_u+\eta_w I_n+\eta_w I_p+ \eta_c I_c)}{N_0},
\end{equation}
across all susceptible pools (see Table~\ref{tab:params} for details\david{there are no details in the table; you need to do a better job of explaining what's going on}), (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S_p(0)=S_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta_w$  & Relative probability of transmission for ``waiting'' individuals & - & - \\ \hline
  $\eta_c$  & Relative probability of transmission for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.39 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $W_S, W_I, W_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by solving the coupled system including $S_u+S_n=N_0$ and $F_S S_u-\omega S_n=0$.
\david{you mean by solving the algebraic equations obtain by setting the DEs to 0.  not clear why you bother to mention two of these algebraic eqns.}
The DFE is
\begin{equation}
\label{dfe}
S_n^*= \frac{\rho}{\omega} N_0, \ S_u^*= N_0-S_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
\david{above you had j in Roman font even though it *is* a variable, so I changed to $j$}

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (A \times S_u^* + B \times S_n^*) \times C, 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
A=& \gamma(\omega+\gamma) + (\gamma \eta_w + \omega \eta_c p_I) F_I, \\
B=& \big(\omega+(F_I+\gamma)\eta_w\big) \gamma+\frac{(\eta_w \gamma+ \eta_c\omega) \omega p_I F_I }{\omega+\gamma}, \\ 
C=& \frac{1}{N_0 (\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I))}.
\end{aligned}
\end{equation}
\david{put $\beta/\gamma$ in front of the expression in eq \eqref{R0}, so it is clear that the rest is a modification of the the simplest expression.  It would be worth spending some time trying to write the expression in a way that helps the reader understand where the different factors come from.}
Further details of derivation of $\Rnum$ are provided in the appendix section.

The analytical calculation of the next generation matrix, $G$\david{don't waste the symbol $G$ on a quantity we don't really need; $G$ is often used for generation interval}, and the derivation of $\Rnum$ was carried out in Maple \citep{maple14} by using simple linear Algebra package. 
  We used R \citep{r} for the simulation part which included using the explicit expression of $\Rnum$ \ref{R0} as a function of the underlying parameters, specifying the parameters a realistic range, calculating the corresponding $\Rnum$ and plotting the contours of $\Rnum$ by using ggplot package (ref?) in R. The related result is presented in Figure \ref{pan}, where  panel \eqref{p.a} represents the random testing, and panel \eqref{p.b} is representing non-random testing. The parameter values or ranges are as follows (I like using a table here, with the param, their units and values).\david{use comment macro for comments} It is notable that in the simulation part, the explicit expression of $\Rnum$ was used and not the approximation \eqref{eq:R0appr}.\david{that's the first mention of ``the approximation''} This simulation reflects the behavior of $\Rnum$ with respect to the selected parameters for two different testing strategies: (i) random testing, represented by all testing weights to be the same, $W_S=W_I=W_R$, and (ii) non-random testing, when testing weight are not equal. For simulation purposes we chose $W_S=W_I=W_R=1$ for random testing, and $W_S=0.3$ and $W_I=W_R=1$ for non-random testing strategy. Note that the critical contour of $\Rnum=1$ is plotted in solid line in Figure \ref{pan}.

\david{Details such as the fact that you used ggplot are not important to state in the main text.  On the other hand, there isn't a clear narrative here.  The reader should feel that a story is being told.}

@@

The explicit formula for the basic reproduction number, $\Rnum$ \eqref{R0}, provides an opportunity to study the influence of changes in the underlying parameters on the critical index of epidemic dynamics. We are interested in understanding the effect of parameters that can be realistically controlled by changing in isolation, \percap testing intensity and test resulting, i.e., $\eta_c$ and $\eta_w$, $\rho$ and $\omega$, respectively. The following Proposition is the direct result of taking partial derivative of $\Rnum$ \eqref{R0} with respect to selected parameters. Note that we used Taylor approximation of $\Rnum$ at $\rho=0$ due to the complexity of the expression and thus inconclusive.\david{in that case it would be good to establish numerically that the relationships given in the proposition are valid more generally, at least for reasonable parameter ranges}
See the appendix for details.

\david{make sure to use $\Rnum$ not $\Rnum$ below (and often in the supp material section).  More importantly, you really need to state in words what this proposition is saying.  The reader has to check the param table and think it through themselves.}
\begin{proposition}
\label{prop1}
Using the expression of $\Rnum$ \eqref{R0},
\begin{enumerate}
\item \label{p1:eta}
$\partial{\Rnum}/\partial{\eta_c} \geq 0$ and $\partial{\Rnum}/\partial{\eta_w} \geq 0$. 
\item \label{p1:rho}
$\partial{\Rnum}/\partial{\rho} \leq 0$ when $\rho \approx 0$.
\item \label{p1:omega}
$\partial{\Rnum}/\partial{\omega}$ can be positive or negative when $\rho \approx 0$.
\end{enumerate}
\end{proposition}

Given that the perfect isolation occurs at $\eta_c=\eta_w = 0$,
\david{That's a perfect example of the kind of phrasing that really must be changed.  You need to say something like ``occurs when no one transmits while waiting for test results (i.e., $\eta_c=\eta_w = 0$)''.  It makes the text so much easier to process and doesn't require the reader to keep looking at the table to make sense of what you're saying.}
Prop.~\ref{p1:eta} means that lifting the isolation from awaiting group results in an increased $\Rnum$ and consequently a greater number of infected individuals. Prop.~\ref{p1:rho} indicates that increasing the \percap testing intensity, $\rho$, reduces $\Rnum$. Lastly, Prop.~\ref{p1:omega} means that speeding up the test reporting may or may not reduce $\Rnum$, it will depend on other parameters. It is straightforward to derive \ref{p1:rho} and \ref{p1:omega} from the Taylor approximation of $\Rnum$ at $\rho=0$.
\david{can we gain some insight from the expressions for the partial derivatives in the proposition, as opposed to just their signs?}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of the behaviour of the basic reproduction number, $\Rnum$, between random versus targeted testing strategies at differnt levels of testing and isolation.}
We numerically evaluate $\Rnum$ \eqref{R0} using the following parameters (listed in Table \ref{tab:params}):
$N_0=1 \times 10^6$, $\omega \in [0.1,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,0.01]$ 1/day, $\eta_w \in [0,1]$, $\eta_c \in [0,1]$, $\beta=0.39$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Rnum$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $W_S=W_I=W_R=1$ and (b) targeted testing where $W_S=0.3$ and $W_I=W_R=1$. The black solid line in each panel represents the critical contour of $\Rnum=1$. \bmb{we still have some collisions in the y-axis tick labels}}
\label{pan}
\end{figure}

Our numerical solutions in support of the analytical results in Proposition \ref{prop1} are presented in Figure \ref{pan}. It can be inferred that when random testing strategy is applied, Figure \ref{p.a}, comparing to the targeted testing strategy, Figure \ref{p.b}, the critical \percap testing intensity of the corresponding panels are lower in non-random testing. Also, it appears that speeding the test reporting, i.e., increasing $\omega$, does not significantly lower $\Rnum$ when awaiting people (including people in $I_p$ and $I_n$ compartments) follow the isolation perfectly, i.e., $\eta_w$ is closer to 0. However, speeding the test reporting reduces the epidemic more when the awaiting people less follow the isolation, i.e., $\eta_w$ is closer to 1.

@@

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $X_n$, $X_p$ and $X_c$ for $X \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta_w=0$ and $\eta_c=0$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Again the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

@@

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf}. To summarise, returning test results more rapidly tends to be favourable (i.e., reduces $\Rnum$) when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive (i.e., $p_I$ is close to 1), and (iii) the targeted testing strategy is used (i.e., $W_i \gg W_s$).

@@

The model in the form of a system of ordinary differential equations is 
\begin{subequations}\label{model}
\begin{align}
 d S_u/dt &= -\Lambda S_u - F_S S_u + \omega S_n, \\
 d S_n/dt &= -\Lambda S_n + (1-p_S) F_S S_u - \omega S_n, \\
 d S_p/dt &= -\Lambda S_p + p_S F_S S_u - \omega S_p, \\
 d S_c/dt &= -\Lambda S_c + \omega S_p, \\
 d I_u/dt &= \Lambda S_u - F_I I_u + \omega I_n  - \gamma I_u, \\
 d I_n/dt &= \Lambda S_n + (1-p_I) F_I I_u - \omega I_n -\gamma I_n, \\
 d I_p/dt &= \Lambda S_p + p_I F_I I_u - \omega I_p -\gamma I_p, \\
 d I_c/dt &= \Lambda S_c + \omega I_p - \gamma I_c, \\
 d R_u/dt &= \gamma I_u - F_R R_u + \omega R_n, \\
 d R_n/dt &= \gamma I_n + (1-p_R) F_R R_u - \omega R_n, \\
 d R_p/dt &= \gamma I_p + p_R F_R R_u  - \omega R_p, \\
 d R_c/dt&= \gamma I_c + \omega R_p, \\
 dN/dt &= \omega (S_n + I_n + R_n),  \\
 dP/dt &= \omega(I_p + R_p) ,
\end{align}
\end{subequations}

>>>>

\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\mathcal{R}_0}
\newcommand{\covid}{COVID-19\xspace}

\newcommand*\subtxt[1]{_{\textnormal{#1}}}
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

\newcommand{\comment}{\showcomment}
\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

@@

We developed a deterministic model, Eqs. \eqref{model}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Symbolically, the testing status of an individual in the disease compartment $X$, where $X \in \{S,I,R\}$ , is reflected in the subscript, namely $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively. Further, two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of the basic reproduction number $\Rnum$ are presented in the appendixs.

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=0.3]{./pix/sir_comp.png}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. For the description of the parameters see \cref{tab:params}. \todo{change the subs to roman font}
\label{fig:flowchart}}
\end{center} 
\end{figure}

Table~\ref{tab:params} defines the model parameters, which are generally straightforward
\percap flows between compartments, or modifiers to these flow rates. The \percap testing intensity is defined as the number of daily tests taken in a population of size $N_0$. 
The novel component of the model comes in through the compartment-specific relative testing weights $W_S$, $W_I$ and $W_R$; these give the relative rates at which people in the $S$, $I$, and $R$ compartments are tested, respectively. For example, $W_I/W_S=2$ means that infected individuals are tested at twice the \percap rate of susceptible individuals.

In order to link to more applied models\david{what does that mean?  what is an ``applied model''?}, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S\_u + W_I I\_u + W_R R\_u$, and calculating a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, the \percap testing rate for compartment $X$ is $F_X=\sigma W_X$, where $X \in \{S,I,R\}$. 
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I\_c$, $R\_c$) compartments and are thus unavailable for further testing.  Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can (if large enough) exhaust the pool of people available for testing, leading to a singularity when no one is left untested.  Although this phenomenon does not affect our analysis of $\Rnum$, it can affect the temporal dynamics (we discuss an adjustment to the model that solves this problem in the appendix).

\david{A comment on notation:  You haven't actually defined $S_u$, $I_u$ etc.  In any case, I urge you to use macros and use Roman font for subscripts that are textual as opposed to variables.  So $S_{\rm u}$ and $I_{\rm p}$ etc.  Then it is clear the subscript isn't a variable with a value.}\ali{done, $p_S$ as is?}

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population\david{What's the difference btw ``well-mixed'' and ``homoegeneous''?}\ali{ non structured eg age, ok?}, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection, $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I\_u+\eta\_w I\_n+\eta\_w I\_p+ \eta\_c I\_c)}{N_0},
\end{equation}
across all susceptible pools (see Table~\ref{tab:params} for details\david{there are no details in the table; you need to do a better job of explaining what's going on}\ali{help?}), (ii) $\eta\_c \leq \eta\_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S\_p(0)=S\_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta\_w$  & Relative probability of transmission for ``waiting'' individuals & - & - \\ \hline
  $\eta\_c$  & Relative probability of transmission for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.39 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $W_S, W_I, W_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by setting the differential equations to 0 and solving for the unknowns (\ali{is this clear or the "unkowns" needs to be defined? also * in sup for I and R?}. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (A \times S\_u^* + B \times S\_n^*) \times C, 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
A=& \gamma(\omega+\gamma) + (\gamma \eta\_w + \omega \eta\_c p_I) F_I, \\
B=& \big(\omega+(F_I+\gamma)\eta\_w\big) \gamma+\frac{(\eta\_w \gamma+ \eta\_c\omega) \omega p_I F_I }{\omega+\gamma}, \\ 
C=& \frac{1}{N_0 (\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I))}.
\end{aligned}
\end{equation}
Further details of derivation of $\Rnum$ are provided in the appendix section.

The analytical calculation of the next generation matrix and the derivation of $\Rnum$ was carried out in Maple \citep{maple14} by using simple linear Algebra package. 
  We used R \citep{r} for the simulation part which included using the explicit expression of $\Rnum$ \ref{R0} as a function of the underlying parameters, specifying the parameters a realistic range, calculating the corresponding $\Rnum$ and plotting the contours of $\Rnum$ \eqref{R0}. The related result is presented in Figure \ref{pan}, where  panel \eqref{p.a} represents the random testing, and panel \eqref{p.b} is representing non-random testing. The parameter values or ranges are presented in Table~\ref{tab:params}. This simulation reflects the behavior of $\Rnum$ with respect to the selected parameters for two different testing strategies: (i) random testing, represented by all testing weights to be the same, $W_S=W_I=W_R$, and (ii) non-random testing, when testing weight are not equal. For simulation purposes we chose $W_S=W_I=W_R=1$ for random testing, and $W_S=0.3$ and $W_I=W_R=1$ for non-random testing strategy. Note that the critical contour of $\Rnum=1$ is plotted in solid line in Figure \ref{pan}.

\david{Details such as the fact that you used ggplot are not important to state in the main text.  On the other hand, there isn't a clear narrative here.  The reader should feel that a story is being told.}\ali{partially addressed}

@@

The explicit formula for the basic reproduction number, $\Rnum$ \eqref{R0}, provides an opportunity to study the influence of changes in the underlying parameters on the critical index of epidemic dynamics. We are interested in understanding the effect of parameters that can be realistically controlled by changing in isolation, \percap testing intensity and test resulting, i.e., $\eta\_c$ and $\eta\_w$, $\rho$ and $\omega$, respectively. The following Proposition is the direct result of taking partial derivative of $\Rnum$ \eqref{R0} with respect to selected parameters. Note that we used Taylor approximation of $\Rnum$ at $\rho=0$ due to the complexity of the expression and thus inconclusive.\david{in that case it would be good to establish numerically that the relationships given in the proposition are valid more generally, at least for reasonable parameter ranges}\ali{help}
See the appendix for details.

\david{More importantly, you really need to state in words what this proposition is saying.  The reader has to check the param table and think it through themselves.}
\begin{proposition}
\label{prop1}
Using the expression of $\Rnum$ \eqref{R0},
\begin{enumerate}
\item \label{p1:eta}
$\partial{\Rnum}/\partial{\eta\_c} \geq 0$ and $\partial{\Rnum}/\partial{\eta\_w} \geq 0$. 
\item \label{p1:rho}
$\partial{\Rnum}/\partial{\rho} \leq 0$ when $\rho \approx 0$.
\item \label{p1:omega}
$\partial{\Rnum}/\partial{\omega}$ can be positive or negative when $\rho \approx 0$.
\end{enumerate}
\end{proposition}

Given that the perfect isolation occurs when no one transmits while waiting for test results (i.e., $\eta\_c=\eta\_w = 0$), Prop.~\ref{p1:eta} means that lifting the isolation from awaiting group results in an increased $\Rnum$ and consequently a greater number of infected individuals. Prop.~\ref{p1:rho} indicates that increasing the \percap testing intensity, $\rho$, reduces $\Rnum$. Lastly, Prop.~\ref{p1:omega} means that speeding up the test reporting may or may not reduce $\Rnum$, it will depend on other parameters. It is straightforward to derive \ref{p1:rho} and \ref{p1:omega} from the Taylor approximation of $\Rnum$ at $\rho=0$.
\david{can we gain some insight from the expressions for the partial derivatives in the proposition, as opposed to just their signs?}\ali{help}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of the behaviour of the basic reproduction number, $\Rnum$, between random versus targeted testing strategies at differnt levels of testing and isolation.}
We numerically evaluate $\Rnum$ \eqref{R0} using the following parameters (listed in Table \ref{tab:params}):
$N_0=1 \times 10^6$, $\omega \in [0.1,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,0.01]$ 1/day, $\eta\_w \in [0,1]$, $\eta\_c \in [0,1]$, $\beta=0.39$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Rnum$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $W_S=W_I=W_R=1$ and (b) targeted testing where $W_S=0.3$ and $W_I=W_R=1$. The black solid line in each panel represents the critical contour of $\Rnum=1$. \bmb{we still have some collisions in the y-axis tick labels}}
\label{pan}
\end{figure}

Our numerical solutions in support of the analytical results in Proposition \ref{prop1} are presented in Figure \ref{pan}. It can be inferred that when random testing strategy is applied, Figure \ref{p.a}, comparing to the targeted testing strategy, Figure \ref{p.b}, the critical \percap testing intensity of the corresponding panels are lower in non-random testing. Also, it appears that speeding the test reporting, i.e., increasing $\omega$, does not significantly lower $\Rnum$ when awaiting people (including people in $I\_p$ and $I\_n$ compartments) follow the isolation perfectly, i.e., $\eta\_w$ is closer to 0. However, speeding the test reporting reduces the epidemic more when the awaiting people less follow the isolation, i.e., $\eta\_w$ is closer to 1.

@@

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $X_n$, $X_p$ and $X_c$ for $X \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta\_w=0$ and $\eta\_c=0$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Again the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

@@

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf}. To summarise, returning test results more rapidly tends to be favourable (i.e., reduces $\Rnum$) when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta\_c \ll \eta\_w$), (ii) the test is highly sensitive (i.e., $p_I$ is close to 1), and (iii) the targeted testing strategy is used (i.e., $W_i \gg W_s$).

@@

The model in the form of a system of ordinary differential equations is 
\begin{subequations}\label{model}
\begin{align}
 d S\_u/dt &= -\Lambda S\_u - F_S S\_u + \omega S\_n, \\
 d S\_n/dt &= -\Lambda S\_n + (1-p_S) F_S S\_u - \omega S\_n, \\
 d S\_p/dt &= -\Lambda S\_p + p_S F_S S\_u - \omega S\_p, \\
 d S\_c/dt &= -\Lambda S\_c + \omega S\_p, \\
 d I\_u/dt &= \Lambda S\_u - F_I I\_u + \omega I\_n  - \gamma I\_u, \\
 d I\_n/dt &= \Lambda S\_n + (1-p_I) F_I I\_u - \omega I\_n -\gamma I\_n, \\
 d I\_p/dt &= \Lambda S\_p + p_I F_I I\_u - \omega I\_p -\gamma I\_p, \\
 d I\_c/dt &= \Lambda S\_c + \omega I\_p - \gamma I\_c, \\
 d R\_u/dt &= \gamma I\_u - F_R R\_u + \omega R\_n, \\
 d R\_n/dt &= \gamma I\_n + (1-p_R) F_R R\_u - \omega R\_n, \\
 d R\_p/dt &= \gamma I\_p + p_R F_R R\_u  - \omega R\_p, \\
 d R\_c/dt&= \gamma I\_c + \omega R\_p, \\
 dN/dt &= \omega (S\_n + I\_n + R\_n),  \\
 dP/dt &= \omega(I\_p + R\_p) ,
\end{align}
\end{subequations}
